How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-12 doc/paper.tex
added in seth comments round 2. made out of sample all about forecasting to make it clearer to audience

\begin{abstract}
  One of the central tenants of conservation paleobiology is that risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. 
  We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
  To address these questions we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, recent change in geographic range, climate state, and the change in climate state. Our best performing model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.
  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) 
  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are. For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}. By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does a species' extinction risk change over its duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}). 
%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years? 
Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

@@

\subsection{Model adequacy}

We are interested in model adequacy into two contexts: in-sample and out-of-sample predictive performance. In-sample means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. Out-of-sample is defined in context below.

We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data). To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

@@

\subsection{Model adequacy}

@@

For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of posterior predictive AUC estimates for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

\subsection{Cross-validation}

Expected out-of-sample predictive performance was estimated using five-fold cross-validation \citep{ESL}. This procedure yields four posterior (predictive) distributions, each corresponding to AUC values calculated from model-based predictions compared to the extinction state of the hold-out data. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find the visual degree of this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the posterior predictive distributions of expected out-of-sample AUC (Fig. \ref{fig:fold_auc}) reveals a similar range in plausible values for all models as the in-sample AUC posterior predictive distributions (Fig. \ref{fig:auc_hist}). Interestingly, the differences between the posterior predictive distributions for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interepting from the median values is incorrect -- the models are effectively indistinguishable in their expected out-of-sample AUC values.

@@

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). For each taxonomic group, the time-series of posterior predictive values for each model are broadly congruent.

In the analysis of the posterior predictive distributions of the in-sample AUC values for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of out-of-sample AUC values -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly rank the extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original). The similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

One of the most striking results of this analysis is that the in-sample and out-of-sample comparisons between our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them generalize to out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of our in-sample results are that AUC estimates for our models models differ significantly between models and that all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

@@

\beginsupplement
\begin{refsection}
\section{Supplementary Material}

@@

\subsubsection{Covariate transforms and standardizations} \label{sec:data_trans}

>>>>

\begin{abstract}
  A tenants of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

  %risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. 
  %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) 
  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

@@

By analyzing extinction and survival data from the fossil record, the hope is this can aide in predicting the extinction risk of extant species -- after all, the present must at some level be a function of the past. Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinates of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

%For example, while a predictor may be ``significant'' when comparing the odds of extinction risk between groups, the practical difference that predictor makes on prediction can be minimal \citep{ARM}. By including the kinds of biological and abiotic predictors that have been shown to affect differences in extinction risk, we can quantify their actual, as opposed to relative, effects on predictive performance.

A related question is if the changes to biotic or abiotic predictors, and not just their values, are similarly important factors for predicting extinction. For example, we know that a species' global geographic range changes over its duration \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}. We also know that a species' geographic range size is a good predictor of differences extinction risk \citep{Payne2007,Jablonski2003,Jablonski2008,Jablonski2006}. This begs the question: how does a species' extinction risk change over its duration? While the phenomenon of species' geographic range change over time has been studied \citep{Foote2007,Liow2010,Liow2007,Kiessling2013}, the potential predictive impact of this change has been under-evaluated (but see \citet{Kiessling2013}). 
%For example, does a species' extinction risk increase if that species decreased in global geographic range size over 1 million years?

%Here, we explicitly estimate the effects of changing geographic range as well as differences in global climate on how well we can predict species extinction. Importantly, the inclusion of these ``historical'' predictors allows us to more fully evaluate the question of how much information about a species' past is necessary or useful when predicting a species' risk of extinction.

@@

\subsection{In-sample and out-of-sample forecasting}

We are interested in model performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.

In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to. This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). These in-sample measures, however, are not necessarily good estimates of the model's ability to predict data from the future. To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}. The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast forecast the fifth fold.

@@

\subsection{In-sample forecasting adequacy}

@@

For many taxon/model combinations there are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 -- AUC value of 0.5 indicates that the model's predictions are no better than random (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

\subsection{Out-of-sample forecasting performance}

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find the visual degree of this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences means that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.

@@

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). The time-series of posterior predictive values for each model are broadly congruent between each taxonomic group.

In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

One of the most striking results of this analysis is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical coarviates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

Other noteworthy aspects of the results of our in-sample forecasts are that although AUC estimates for our models models differ significantly between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). For our in-sample results, while the statistically best model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance compared to the other models is virtually negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a \textit{statistically} greater AUC than the other three models, this result is not practically or \textit{scientifically} significant.

@@

\beginsupplement
\begin{refsection}

%\section{Supplemental Figures}

\section{Supplement to Materials and Methods}

@@

\subsubsection{Covariate transforms and standardizations}
